The dynamics of starvation and recovery

### 2017-03-09 Manuscript/Revision_2017/17.tex
some ms changes

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals.
Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments.
To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality.
As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state.
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships between body size and a variety of traits and find that population dynamics subject to these constraints are typically driven to a steady state.
Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized.
Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations.
Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percentage of body fat.
For body masses $\lesssim 10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $\gtrsim 10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

{\color{red} Consumers transition from the full state $F$ to the hungry state $H$ at a rate $\sigma$---the starvation rate---and also in proportion to the absence of resources $(1-R)$  (see SI for detailed a description and derivation of all rates and dependencies).  Conversely, consumers recover from state $H$ to state $F$ at rate $\xi \rho$ and in proportion to $R$, where $\xi$ represents a ratio between maximal resource consumption and the carrying capacity of the resource. %and $\approx 2$.
Resources are eaten by the hungry consumers at rate $\rho R + \delta$, that accounts for their somatic growth ($\rho R$) and  maintenance ($\delta$). Full consumers eat resources at a constant rate $\beta$ that accounts for maximal maintenance and somatic growth (please see SI for a mechanistic derivation of these terms from resource energetics).}
%This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.
The NSM represents an ecologically motivated fundamental extension of the idealized starving random walk model of foraging, which focuses on resource depletion, to include reproduction and resource replenishment~\cite{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}, and is a more general formulation than previous models incorporating starvation~\cite{Persson:1998hz}.

In the mean-field approximation, in which the consumers and resources are perfectly mixed, their densities evolve according to the rate equations
\begin{align}
\label{eq:system}
\begin{split}
\frac{dF}{dt} &= \lambda F + \xi \rho RH - \sigma \left(1-R\right)F,  \\
\frac{dH}{dt} &= \sigma \left(1-R\right)F - \xi \rho RH - \mu H,  \\
\frac{dR}{dt} &= \alpha R\left(1-R\right) -\left(\rho R+\delta\right)H-\beta F
\end{split}
\end{align}
This system of nondimensional equations follows from a set of first-principle relationships for resource consumption and growth (see SI for a full derivation and the dimensional form).
Notice that the total consumer density $F+H$ evolves according to $\frac{dF}{dt}+\frac{dH}{dt}=\lambda F-\mu H$.
This resembles the equation of motion for the predator density in the classic Lotka-Volterra model~\cite{murray2011mathematical}, except that the resource density does not appear in the growth term.
As discussed above, the attributes of reproduction and mortality have been explicitly apportioned to the full and hungry consumers, respectively, so that the growth in the total density is decoupled from the resource density.

Equation~\eqref{eq:system} has three fixed points: two trivial fixed points at $(F^*,H^*,R^*)=(0,0,0)$ and $(0,0,1)$, and one non-trivial, internal fixed point at
\begin{align}
\label{eq:ss}
\begin{split}
F^* &= \frac{(\sigma-\lambda) \alpha  \lambda  \mu ^2  (\mu +\xi  \rho )}{A (\lambda  \rho  B+\mu  \sigma  (\beta  \mu +\lambda  (\delta +\rho )))}, \\
H^* &= \frac{(\sigma-\lambda) \alpha  \lambda ^2 \mu  (\mu +\xi  \rho )}{A (\lambda  \rho  B+\mu  \sigma  (\beta  \mu +\lambda  (\delta +\rho )))}, \\
R^* &= (\sigma - \lambda)\frac{\mu  }{A}.
\end{split}
\end{align}
where $A=(\lambda  \xi  \rho +\mu  \sigma )$ and $B=(\beta  \mu  \xi +\delta  \lambda  \xi -\lambda  \mu )$. The stability of this fixed point is determined by the Jacobian matrix $\bf J$, where each matrix element $J_{ij}=\partial{\dot X_i}/\partial{X_j}$ when evaluated at the internal fixed point, and $\mathbf{X}$ is the vector $(F,H,R)$.
The parameters in Eq.~\eqref{eq:system} are such that the real part of the largest eigenvalue of $\mathbf{J}$ is negative, so that the system is stable with respect to small perturbations from the fixed point.
Because this fixed point is unique, it is the global attractor for all population trajectories for any initial condition where the resource and consumer densities are both nonzero.

From Eq.~\eqref{eq:ss}, an obvious constraint on the NSM is that the reproduction rate $\lambda$ must be less than the starvation rate $\sigma$, so that $R^*$ is positive.  In fact, when the resource density $R=0$, the rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ represents a transcritical (TC) bifurcation~\cite{Strogatz:2008wo} that demarcates a physical  from an unphysical regime where all densities become less than zero.
The biological implication of the constraint $\lambda<\sigma$ has a simple interpretation---the rate at which a macroscopic organism loses mass due to lack of resources is generally much faster than the rate of reproduction.
As we will discuss below, this inequality is a natural consequence of allometric constraints~\cite{Kempes:2012hy} for organisms within empirically observed body size ranges. %(Fig.~\ref{fig:gvs})

In the physical regime of $\lambda<\sigma$, the fixed point \eqref{eq:ss} may either be a stable node or a limit cycle (Fig.~\ref{fig:fp}).
In continuous-time systems, a limit cycle arises when a pair of complex conjugate eigenvalues crosses the imaginary axis to attain positive real parts~\cite{GuckHolmes}.
This Hopf bifurcation is defined by ${\rm Det}({\bf S}) = 0$, with $\bf S$ the Sylvester matrix, which is composed of the coefficients of the characteristic polynomial of the Jacobian matrix~\cite{Gross:2004p2428}.
As the system parameters are tuned to be within the stable regime, but close to the Hopf bifurcation, the amplitude of the transient cycles become large.
Given that ecological systems are constantly being perturbed~\cite{Hastings:2001jh}, the onset of transient cycles, even though they decay with time in the mean-field description, can increase the extinction risk~\cite{Neubert:1997wk,Caswell:2005eo,Neubert:2009td}.
%Thus the distance of a system from the Hopf bifurcation provides a measure of its persistence.

When the starvation rate $\sigma\gg\lambda$, a substantial fraction of the consumers are driven to the hungry non-reproducing state.
Because reproduction is inhibited, there is a low steady-state consumer density and a high steady-state resource density.
However, if $\sigma/\lambda\to 1$ from above, the population is overloaded with energetically-replete (reproducing) individuals, thereby promoting oscillations between the consumer and resource densities (Fig.~\ref{fig:fp}).

Whereas the relation between consumer growth rate $\lambda$ and the starvation rate $\sigma$ defines an absolute bound of biological feasibility---the TC bifurcation---$\sigma$ also determines the sensitivity of the consumer population to changes in resource density.
When $\sigma\gg\lambda$, the steady-state population density is small, thereby increasing the risk of stochastic extinction.
On the other hand, as $\sigma$ decreases, the system will ultimately be poised either near the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).
If the recovery rate $\rho$ is sufficiently small, the TC bifurcation is reached and the resource eventually is eliminated.
If $\rho$ exceeds a threshold value, cyclic dynamics will develop as the Hopf bifurcation is approached.
\\

%In the case of the 2-stage consumer model, the Jacobian evaluated at the internal steady state (denoted by $|_*$) is written
%
%\begin{equation}
%\mathbf{J}|_* =
%%\left(
%%\begin{array}{ccc}
%% -F^* m+(1-R^*) a -R^* a -H^* \rho  & -R^* \rho  & -m R^* \\
%% -H^* \rho -F^* \sigma  & -\mu -R^* \rho  & (1-R^*) \sigma  \\
%% H^* \rho +F^* \sigma  & R^* \rho  & \lambda -(1-R^*) \sigma  \\
%%\end{array}
%%\right).
%\left(
%\begin{array}{ccc}
% -\frac{\lambda  \rho  (\sigma - \lambda )}{\lambda  \rho +\mu  \sigma } & \frac{\mu  \rho  (\sigma -\lambda )}{\lambda  \rho +\mu  \sigma } & \frac{\alpha  \lambda  (\mu +\rho )}{m \mu +\lambda  \rho } \\
% \frac{\lambda  (\mu +\rho ) \sigma }{\lambda  \rho +\mu  \sigma } & -\frac{\mu  (\mu +\rho ) \sigma }{\lambda  \rho +\mu  \sigma } & -\frac{\alpha  \lambda  (\mu +\rho )}{m \mu +\lambda  \rho } \\
% -\frac{m \mu  (\sigma - \lambda)}{\lambda  \rho +\mu  \sigma } & -\frac{\mu  \rho  (\sigma - \lambda )}{\lambda  \rho +\mu  \sigma } & -\frac{\alpha  \mu  (\sigma - \lambda)}{\lambda  \rho +\mu  \sigma } \\
%\end{array}
%\right).
%\end{equation}

%CHRIS - we need to define beta in this section...

\noindent {\bf Role of allometry} \\
%[Link Allometry stuff to our model - what does it provide to the story]
%[Introduce biological and linked constraints]
%[Allometries capture vast amounts of diversity via a single parameter --- body size]
%[Allometries have captured intterest etc across multiple scales and biological classes of organisms]
While there are no a priori constraints on the parameters in the NSM, most
organisms correspond to restricted portions of the parameter space.  Here we
use allometric scaling relations to constrain the covariation of rates in a
principled and biologically meaningful manner.  Allometric scaling relations
highlight common constraints and average trends across large ranges in body
size and species diversity. Many of these relations can be derived from a
small set of assumptions and below we describe our framework to determine the
covariation of timescales and rates across the range of mammals for each of
the key parameters of our model (cf.~\cite{Yodzis:1992hg}).  We are thereby
able to define the regime of dynamics occupied by the entire class of mammals, along with the key differences between the largest and smallest mammals.

Nearly all of the rates described in the NSM are determined by consumer
metabolism, which can be used to describe a variety of organismal features
\cite{Brown:2004wq}.  The scaling relation between an organism's metabolic
rate $B$ and its body mass $M$ at reproductive maturity is known to scale as
$B = B_0 M^\eta$~\cite{West:2002it}, where the scaling exponent $\eta$ is
typically close to $2/3$ or $3/4$ for metazoans (e.g., \cite{Brown:2004wq}),
and has taxonomic shifts for unicellular species between $\eta\approx 1$ in
eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}.
%Justin adding an explanation for beta

Several efforts have shown how a partitioning of $B$ between growth and
maintenance purposes can be used to derive a general equation for both the
growth trajectories and growth rates of organisms ranging from bacteria to
metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
of the organism at any point in its development.  This balance has the
general solution \cite{bettencourt,Kempes:2012hy}
\begin{eqnarray}
\label{m1}
\left(\frac{m\left(t\right)}{M}\right)^{1-\eta}\!=1\!-\!\left[1\!-\!\left(\frac{m_{0}}{M}\right)^{1\!-\!\eta}\right]e^{-a\left(1\!-\!\eta\right)t/M^{1-\eta}},
\end{eqnarray}
where, for $\eta<1$, $M=(B_{0}/B_{m})^{1/(1-\eta)}$ is the asymptotic mass, $a=B_{0}/E_{m}$, and $m_0$ is mass at birth, itself varying allometrically (see SI).  We now use this solution to define the timescale for reproduction and recovery from starvation (Fig.~\ref{fig:growth}; see \cite{moses2008rmo} for a detailed presentation of these timescales). The time that it takes to reach a particular mass $\epsilon M$ is given by the timescale
\begin{equation}
\label{t1}
\tau\left(\epsilon\right) = \ln\left[\frac{1-\left(m_{0}/M\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a\left(1-\eta\right)},
\end{equation}
where we will define values of $\epsilon$ to describe a set of rates within our model. For the time to reproduce, $t_{\lambda}=\tau\left(\epsilon_{\lambda}\right)$, where $\epsilon_{\lambda}$ is the fraction of the asymptotic mass where an organism is reproductively mature and should be close to one (typically $\epsilon_{\lambda}\approx0.95$ \cite{West:2001bv}). The growth rate is then given by $\lambda=\ln\left(\upsilon\right)/t_{\lambda}$ where $\upsilon$ is the number of offspring produced, and for any constant value of $\epsilon_{\lambda}$, this rate will scale as $\lambda\propto M^{\eta-1}$ for $M\gg m_{0}$ \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}.

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since
only certain tissues can be digested for energy (for example the brain cannot
be degraded to fuel metabolism), we define the rates for starvation, death,
and recovery by the timescales required to reach, or return from, specific
fractions of the replete-state mass (Fig. \ref{fig:gvs}; see SI, Table I for parameterizations).  We define
$m_{\sigma}=\epsilon_{\sigma} M$, where $\epsilon_{\sigma}<1$ is the fraction of
replete-state mass where reproduction ceases. This fraction will deviate from a constant
if tissue composition systematically scales with adult mass.  For example,
making use of the observation that body fat in mammals scales with overall
body size according to $M_{\rm fat}=f_{0}M^{\gamma}$ and assuming that once
this mass is fully digested the organism starves, this would imply that
$\epsilon_{\sigma}=1-f_{0}M^{\gamma}/M$. It follows that the
recovery timescale, $t_{\rho}$, is the time to go from
$m=\epsilon_{\sigma} \epsilon_{\lambda} M$ to $m=\epsilon_{\lambda}M$ (Fig. \ref{fig:growth}). Using Eqs.~\eqref{m1} and \eqref{t1} this timescale is given by simply considering an adjusted starting mass of $m_{0}^{\prime}=\epsilon_{\sigma}\epsilon_{\lambda}M$, in which case
\begin{eqnarray}
\label{rhotimescale}
t_{\rho}=\ln\left[\frac{1-\left(\epsilon_{\sigma}\epsilon_{\lambda}\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a^{\prime}\left(1-\eta\right)}
\end{eqnarray}
where $a^{\prime}=B_{0}/E_{m}^{\prime}$ accounts for possible deviations in the biosynthetic energetics during recovery (see SI). It should be noted that more complicated ontogenetic models explicitly handle
storage \cite{hou}, whereas this feature is implicitly covered by the body
fat scaling in our framework.

@@

%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although the rate equations \eqref{eq:system} are general, here we focus on
parameterizations for terrestrial-bound endotherms, specifically mammals,
which range from a minimum of $M\approx1$g (the Etruscan shrew
\emph{Suncus etruscus}) to a maximum of $M\approx10^7$g (the late Eocene
to early Miocene Indricotheriinae).  Investigating other classes of organisms
would simply involve altering the metabolic exponents and scalings associate
with $\epsilon$. Moreover, we emphasize that our allometric equations
describe mean relationships, and do not account for the (sometimes
considerable) variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

To examine whether the NSM could provide such a mechanism, we begin by noting that a theoretical upper bound on mammalian body size is given by $\epsilon_\sigma=0$, where mammals are entirely composed of metabolic reserves, and this occurs at $M=8.3\times 10^8$, or $120$ times the mass of a male African elephant.
Next we examine to what extent a more realistic upper bound to body mass may serve as an evolutionary attractor, thus providing a suitable within-lineage mechanism for Cope's rule.{\color{red}
We directly assess the susceptibility of an otherwise homogeneous population to invasion by a mutated subset of the population (denoted by $^\prime$) where individuals have a modified proportion of body fat $M^\prime=M(1+\chi)$. For the allowable values of $\chi$ the adjusted mass should exceed the amount of body fat, $1+\chi>\epsilon_{\sigma}$, and the adjusted time to reproduce must be positive, which given Equation \ref{t1}, implies that $1-\epsilon_{\lambda}^{1-\eta}\left(1+\chi\right)^{1-\eta}>0$. Together these conditions imply that  $\chi\in(-f_0M^{\gamma-1},1/\epsilon_{\lambda}-1)$ where the upper bound approximately equals $0.05$. }The modified mass adjusts our model via the altered rates of starvation $\sigma(M^\prime)$, recovery $\rho(M^\prime)$, and the maintenance of both starving $\delta(M^\prime)$ and full consumers $\beta(M^\prime)$.
Importantly, $\epsilon_\sigma$, which determines the point along the growth curve that defines the state of starved foragers, is assumed to remain unchanged for the invader population (see SI for detailed derivations of invader rates).
%There is no internal fixed point corresponding to a state where both original residents and invaders coexist (except for the trivial state $\chi=0$).

>>>>

\begin{abstract} %250 words
The eco-evolutionary dynamics of species are fundamentally linked to the energetic constraints of its constituent individuals.
Of particular importance are the tradeoffs between reproduction and the dynamics of starvation and recovery in resource-limited environments.
To elucidate the consequences of this tradeoff, we introduce a minimal nutritional state-structured model that incorporates two classes of consumer: nutritionally replete consumers that reproduce, and undernourished, non-reproducing consumers that are susceptible to mortality.
As a function of the transition rates between these two states that are determined by the abundance of resources, the consumer populations can either undergo cyclic dynamics or reach a steady state.
We obtain strong constraints on starvation and recovery rates by deriving allometric scaling relationships between body size and a variety of traits and find that population dynamics subject to these constraints are typically driven to a steady state.
Moreover, we find that these rates fall within a `refuge' in parameter space, where the probability of extinction of the consumer population is minimized.
Thus we identify a potential mechanism that may both drive and constrain the dynamics of animal populations.
Our model provides a natural framework that predicts maximum body size for mammals by determining the relative stability of an otherwise homogeneous population to a mutant population with altered percentage of body fat.
For body masses $< 1.75\times10^7$g, individuals with increased energetic reserves can invade resident populations, and vice versa for body mass $> 1.75\times10^7$g, thus providing a principled mechanism for a within-lineage driver of Cope's rule.
\end{abstract}
\keywords{foraging | starvation|reproduction}

@@

Consumers transition from the full state $F$ to the hungry state $H$ at a rate $\sigma$---the starvation rate---and also in proportion to the absence of resources $(1-R)$.  Conversely, consumers recover from state $H$ to state $F$ at rate $\xi \rho$ and in proportion to $R$, where $\xi$ represents a ratio between maximal resource consumption and the carrying capacity of the resource. %and $\approx 2$.
Resources are eaten by the hungry consumers at rate $\rho R + \delta$, that accounts for their somatic growth ($\rho R$) and  maintenance ($\delta$). Full consumers eat resources at a constant rate $\beta$ that accounts for maximal maintenance and somatic growth (see SI for mechanistic derivations of these rates from resource energetics).
%This inequality accounts for hungry consumers requiring more resources to replace lost body tissue.
The NSM represents an ecologically motivated fundamental extension of the idealized starving random walk model of foraging, which focuses on resource depletion, to include reproduction and resource replenishment~\cite{Benichou:2014wu,Benichou:2016wl,Chupeau:2016jf}, and is a more general formulation than previous models incorporating starvation~\cite{Persson:1998hz}.

In the mean-field approximation, in which the consumers and resources are perfectly mixed, their densities evolve according to the rate equations
\begin{align}
\label{eq:system}
\begin{split}
\frac{dF}{dt} &= \lambda F + \xi \rho RH - \sigma \left(1-R\right)F,  \\
\frac{dH}{dt} &= \sigma \left(1-R\right)F - \xi \rho RH - \mu H,  \\
\frac{dR}{dt} &= \alpha \left(1-R\right)R -\left(\rho R+\delta\right)H-\beta F
\end{split}
\end{align}
This system of nondimensional equations follows from a set of first-principle relationships for resource consumption and growth (see SI for a full derivation and the dimensional form).
Notice that the total consumer density $F+H$ evolves according to $\frac{dF}{dt}+\frac{dH}{dt}=\lambda F-\mu H$.
This resembles the equation of motion for the predator density in the classic Lotka-Volterra model~\cite{murray2011mathematical}, except that the resource density does not appear in the growth term.
As discussed above, the attributes of reproduction and mortality have been explicitly apportioned to the full and hungry consumers, respectively, so that the growth in the total density is decoupled from the resource density.

Equation~\eqref{eq:system} has three fixed points: two trivial fixed points at $(F^*,H^*,R^*)=(0,0,0)$ and $(0,0,1)$, and one non-trivial, internal fixed point at
\begin{align}
\label{eq:ss}
\begin{split}
F^* &= (\sigma-\lambda)\frac{ \alpha  \lambda  \mu ^2  (\mu +\xi  \rho )}{A (\lambda  \rho  B+\mu  \sigma  (\beta  \mu +\lambda  (\delta +\rho )))}, \\
H^* &= (\sigma-\lambda)\frac{ \alpha  \lambda ^2 \mu  (\mu +\xi  \rho )}{A (\lambda  \rho  B+\mu  \sigma  (\beta  \mu +\lambda  (\delta +\rho )))}, \\
R^* &= (\sigma - \lambda)\frac{\mu  }{A}.
\end{split}
\end{align}
where $A=(\lambda  \xi  \rho +\mu  \sigma )$ and $B=(\beta  \mu  \xi +\delta  \lambda  \xi -\lambda  \mu )$. The stability of this fixed point is determined by the Jacobian matrix $\bf J$, where each matrix element $J_{ij}=\partial{\dot X_i}/\partial{X_j}$ when evaluated at the internal fixed point, and $\mathbf{X}$ is the vector $(F,H,R)$.
The parameters in Eq.~\eqref{eq:system} are such that the real part of the largest eigenvalue of $\mathbf{J}$ is negative, so that the system is stable with respect to small perturbations from the fixed point.
Because this fixed point is unique, it is the global attractor for all population trajectories for any initial condition where the resource and consumer densities are both nonzero.

From Eq.~\eqref{eq:ss}, an obvious constraint on the NSM is that the reproduction rate $\lambda$ must be less than the starvation rate $\sigma$, so that the consumer and resource densities are positive.
%In fact, when the resource density $R=0$, the rate equation for $F$ gives exponential growth of $F$ for $\lambda>\sigma$.
The condition $\sigma = \lambda$ thus represents a transcritical (TC) bifurcation~\cite{Strogatz:2008wo} that demarcates a physical from an unphysical regime where all steady-state densities become negative after intersecting the trivial fixed point $(F^*,H^*,R^*)=(0,0,0)$.
The biological implication of the constraint $\lambda<\sigma$ has a simple interpretation---the rate at which a macroscopic organism loses mass due to lack of resources is generally much faster than the rate of reproduction.
As we will discuss below, this inequality is a natural consequence of allometric constraints~\cite{Kempes:2012hy} for organisms within empirically observed body size ranges. %(Fig.~\ref{fig:gvs})

In the physical regime of $\lambda<\sigma$, the fixed point \eqref{eq:ss} may either be a stable node or a limit cycle (Fig.~\ref{fig:fp}).
In continuous-time systems, a limit cycle arises when a pair of complex conjugate eigenvalues crosses the imaginary axis to attain positive real parts~\cite{GuckHolmes}.
This Hopf bifurcation is defined by ${\rm Det}({\bf S}) = 0$, with $\bf S$ the Sylvester matrix, which is composed of the coefficients of the characteristic polynomial of the Jacobian matrix~\cite{Gross:2004p2428}.
As the system parameters are tuned to be within the stable regime, but close to the Hopf bifurcation, the amplitude of the transient cycles becomes large.
Given that ecological systems are constantly being perturbed~\cite{Hastings:2001jh}, the onset of transient cycles, even though they decay with time in the mean-field description, can increase the extinction risk~\cite{Neubert:1997wk,Caswell:2005eo,Neubert:2009td}.
%Thus the distance of a system from the Hopf bifurcation provides a measure of its persistence.

When the starvation rate $\sigma\gg\lambda$, a substantial fraction of the consumers are driven to the hungry non-reproducing state.
Because reproduction is inhibited, there is a low steady-state consumer density and a high steady-state resource density.
However, if $\sigma/\lambda\to 1$ from above, the population is overloaded with energetically-replete (reproducing) individuals, thereby promoting transient oscillations between the consumer and resource densities (Fig.~\ref{fig:fp}).
If the starvation rate is low enough that the Hopf bifurcation is crossed, these oscillations become stable over time.
This threshold occurs at higher values of the starvation rate as the recovery rate $\rho$ increases, such that the range of parameter space giving rise to cyclic dynamics also increases with higher recovery rates.
\\
% Whereas the relation between consumer growth rate $\lambda$ and the starvation rate $\sigma$ defines an absolute bound of biological feasibility---the TC bifurcation---$\sigma$ also determines the sensitivity of the consumer population to changes in resource density.
% When $\sigma\gg\lambda$, the steady-state population density is small, thereby increasing the risk of stochastic extinction.
% On the other hand, as $\sigma$ decreases, the system will ultimately be poised either near the TC or the Hopf bifurcation (Fig.~\ref{fig:fp}).
% If the recovery rate $\rho$ is sufficiently small, the TC bifurcation is reached and the resource eventually is eliminated.
% If $\rho$ exceeds a threshold value, cyclic dynamics will develop as the Hopf bifurcation is approached.

\noindent {\bf Role of allometry} \\
%[Link Allometry stuff to our model - what does it provide to the story]
%[Introduce biological and linked constraints]
%[Allometries capture vast amounts of diversity via a single parameter --- body size]
%[Allometries have captured intterest etc across multiple scales and biological classes of organisms]
While there are no a priori constraints on the parameters in the NSM, most
organisms correspond to restricted portions of the parameter space.  Here we
use allometric scaling relations to constrain the covariation of rates in a
principled and biologically meaningful manner.  Allometric scaling relations
highlight common constraints and average trends across large ranges in body
size and species diversity. Many of these relations can be derived from a
small set of assumptions and below we describe our framework to determine the
covariation of timescales and rates across the range of mammals for each of
the key parameters of our model (cf. ref.~\citenum{Yodzis:1992hg}).  We are thereby
able to define the regime of dynamics occupied by the entire class of mammals, along with the key differences between the largest and smallest mammals.

Nearly all of the rates described in the NSM are determined by consumer
metabolism, which can be used to describe a variety of organismal features
\cite{Brown:2004wq}.  The scaling relation between an organism's metabolic
rate $B$ and its body mass $M$ at reproductive maturity is known to scale as
$B = B_0 M^\eta$~\cite{West:2002it}, where the scaling exponent $\eta$ is
typically close to $2/3$ or $3/4$ for metazoans (e.g., ref. \citenum{Brown:2004wq}),
and has taxonomic shifts for unicellular species between $\eta\approx 1$ in
eukaryotes and $\eta\approx 1.76$ in bacteria
\cite{DeLong:2010dy,Kempes:2012hy}.
%Justin adding an explanation for beta

Several efforts have shown how a partitioning of $B$ between growth and
maintenance purposes can be used to derive a general equation for both the
growth trajectories and growth rates of organisms ranging from bacteria to
metazoans
\cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}. This relation is derived from the simple balance condition \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}
\begin{eqnarray}
\label{balance}
B_{0}m^{\eta}=E_{m}\frac{dm}{dt}+B_{m}m\,,
\end{eqnarray}
where $E_{m}$ is the energy needed to synthesize a unit of mass, $B_{m}$ is
the metabolic rate to support an existing unit of mass, and $m$ is the mass
of the organism at any point in its development.  This balance has the
general solution \cite{bettencourt,Kempes:2012hy}
\begin{eqnarray}
\label{m1}
\left(\frac{m\left(t\right)}{M}\right)^{1-\eta}\!=1\!-\!\left[1\!-\!\left(\frac{m_{0}}{M}\right)^{1\!-\!\eta}\right]e^{-a\left(1\!-\!\eta\right)t/M^{1-\eta}},
\end{eqnarray}
where, for $\eta<1$, $M=(B_{0}/B_{m})^{1/(1-\eta)}$ is the asymptotic mass, $a=B_{0}/E_{m}$, and $m_0$ is mass at birth, itself varying allometrically (see SI).  We now use this solution to define the timescale for reproduction and recovery from starvation (Fig.~\ref{fig:growth}; see \cite{moses2008rmo} for a detailed presentation of these timescales). The time that it takes to reach a particular mass $\epsilon M$ is given by the timescale
\begin{equation}
\label{t1}
\tau\left(\epsilon\right) = \ln\left[\frac{1-\left(m_{0}/M\right)^{1-\eta}}{1-\epsilon^{1-\eta}}\right]\frac{M^{1-\eta}}{a\left(1-\eta\right)},
\end{equation}
where we will define values of $\epsilon$ to describe a set of rates within our model. For the time to reproduce, $t_{\lambda}=\tau\left(\epsilon_{\lambda}\right)$, where $\epsilon_{\lambda}$ is the fraction of the asymptotic mass where an organism is reproductively mature and should be close to one (typically $\epsilon_{\lambda}\approx0.95$; \citenum{West:2001bv}). The growth rate is then given by $\lambda=\ln\left(\upsilon\right)/t_{\lambda}$ where $\upsilon$ is the number of offspring produced, and for any constant value of $\epsilon_{\lambda}$, this rate will scale as $\lambda\propto M^{\eta-1}$ for $M\gg m_{0}$ \cite{West:2001bv,moses2008rmo,gillooly2002esa,hou,Kempes:2012hy}.

The rate of recovery $\rho = 1/t_\rho$ requires that an organism accrues
sufficient tissue to transition from the hungry to the full state.  Since
only certain tissues can be digested for energy (for example the brain cannot
be degraded to fuel metabolism), we define the rates for starvation, death,
and recovery by the timescales required to reach, or return from, specific
fractions of the replete-state mass (Fig. \ref{fig:gvs}; see SI, Table I for parameterizations).  We define
$m_{\sigma}=\epsilon_{\sigma} M$, where $\epsilon_{\sigma}<1$ is the fraction of
replete-state mass where reproduction ceases. This fraction will deviate from a constant
if tissue composition systematically scales with adult mass.  For example,
making use of the observation that body fat in mammals scales with overall
body size according to $M_{\rm fat}=f_{0}M^{\gamma}$ and assuming that once
this mass is fully digested the organism starves, this would imply that
$\epsilon_{\sigma}=1-f_{0}M^{\gamma}/M$. It follows that the
recovery timescale, $t_{\rho}$, is the time to go from
$m=\epsilon_{\sigma} \epsilon_{\lambda} M$ to $m=\epsilon_{\lambda}M$ (Fig. \ref{fig:growth}). Using Eqs.~\eqref{m1} and \eqref{t1} this timescale is given by simply considering an adjusted starting mass of $m_{0}^{\prime}=\epsilon_{\sigma}\epsilon_{\lambda}M$, in which case
\begin{eqnarray}
\label{rhotimescale}
t_{\rho}=\ln\left[\frac{1-\left(\epsilon_{\sigma}\epsilon_{\lambda}\right)^{1-\eta}}{1-\epsilon_\lambda^{1-\eta}}\right]\frac{M^{1-\eta}}{a^{\prime}\left(1-\eta\right)}
\end{eqnarray}
where $a^{\prime}=B_{0}/E_{m}^{\prime}$ accounts for possible deviations in the biosynthetic energetics during recovery (see SI). It should be noted that more complicated ontogenetic models explicitly handle
storage \cite{hou}, whereas this feature is implicitly covered by the body
fat scaling in our framework.

@@

%\left(\frac{a}{b}\right)^{\frac{1}{\eta -1}}
%where $c=(a/b)^{1/(\eta-1)}$.
Although the rate equations \eqref{eq:system} are general, here we focus on parameterizations for terrestrial-bound endotherms, specifically mammals, which range from a minimum of $M\approx1$g (the Etruscan shrew \emph{Suncus etruscus}) to a maximum of $M\approx10^7$g (the early Oligocene Indricotheriinae and the Miocene Deinotheriinae).  
Investigating other classes of organisms would simply involve altering the metabolic exponents and scalings associate with $\epsilon$. Moreover, we emphasize that our allometric equations describe mean relationships, and do not account for the (sometimes considerable) variance associated with individual species.
\\
%, though this is beyond the current scope of our analysis. %, and we henceforth focus our examination on mammalian species.

@@

To examine whether the NSM could provide such a mechanism, we begin by noting that a theoretical upper bound on mammalian body size is given by $\epsilon_\sigma=0$, where mammals are entirely composed of metabolic reserves, and this occurs at $M=8.3\times 10^8$, or $120$ times the mass of a male African elephant.
Next we examine to what extent a more realistic upper bound to body mass may serve as an evolutionary attractor, thus providing a suitable within-lineage mechanism for Cope's rule.
We directly assess the susceptibility of an otherwise homogeneous population to invasion by a mutated subset of the population (denoted by $^\prime$) where individuals have a modified proportion of body fat $M^\prime=M(1+\chi)$.
For the allowable values of $\chi$ the adjusted mass should exceed the amount of body fat, $1+\chi>\epsilon_{\sigma}$, and the adjusted time to reproduce must be positive, which given Equation \ref{t1}, implies that $1-\epsilon_{\lambda}^{1-\eta}\left(1+\chi\right)^{1-\eta}>0$.
Together these conditions imply that  $\chi\in(-f_0M^{\gamma-1},1/\epsilon_{\lambda}-1)$ where the upper bound approximately equals $0.05$.
The modified mass adjusts our model via the altered rates of starvation $\sigma(M^\prime)$, recovery $\rho(M^\prime)$, and the maintenance of both starving $\delta(M^\prime)$ and full consumers $\beta(M^\prime)$.
Importantly, $\epsilon_\sigma$, which determines the point along the growth curve that defines the state of starved foragers, is assumed to remain unchanged for the invader population (see SI for detailed derivations of invader rates).
%There is no internal fixed point corresponding to a state where both original residents and invaders coexist (except for the trivial state $\chi=0$).
